\chapter{Acknowledgments}

I am sincerely grateful to my advisor, Frank Wood, for his continuous support and guidance from my undergraduate studies through to this thesis. His persistent encouragement toward more ambitious research has shaped me into the researcher I am today and helped me build a body of work of which I am immensely proud.

I would next like to thank my thesis committee members, Kwang Moo Yi and Leonid Sigal, for consistently constructive feedback on my thesis as it turned from a jumble of thoughts and paper excerpts into a coherent document. I'd also like to thank David Poole and Alexandre Bouchard-Côté for their guidance during the earlier iterations of my PhD research.

I'm also indebted to the wonderful collaborators I've worked with throughout my PhD, including Adam Ścibior, Andreas Munk, Andrew Campbell, Andrew Warrington, Alexander Bergholm, Atılım Güneş Baydin, Berend Zwartsenberg, Boyan Beronov, Christian Weilbach, Dylan Green, Hamed Shirzad, Matthew Niedoba, Michael Teng, Saeid Naderiparizi, and Vaden Masrani. I'd also like to thank other current and former members of the lab, my friends and collegaues Abby Azari, Alexander Mead, Ali Nasseri, Charlie He, Jason Yoo, Jonathan Wilder Lavington, Justice Sefas, Ke Zhang, Reuben Cohn-Gordon, Ryan Fayyazi, Vasileios Lioutas, Xiaoxuan Liang. Thank you for the countless hours of academic discussions; of time spent working together; of short breaks from research; and all our other shared memories from my time at UBC. Thank you also to our lab manager Andrew Smith for all of his help during the PhD, and especially for his assistance with navigating the adminstrative hurdles as I wrote up my thesis and completed the PhD.  Thank you especially to Andreas Munk, my roommate for most of this PhD, who helped keep me sane throughout.

Finally, I'd like to thank my wonderful girlfriend, Alex Hoang, for her unwavering understanding and support as I navigated the challenges of this journey.