%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = diss.tex

\chapter{Preface}

The main contributions of this dissertation have been published in the following, with William Harvey as first author:
\begin{enumerate}
    \item \bibentry{harvey2023visual} \label{harvey2023visual}
    \item \bibentry{harvey2022flexible} \label{harvey2022flexible}
    \item \bibentry{harvey2021conditional} \label{harvey2021conditional}
\end{enumerate}
In particular, some of \cref{ch:conditional-diffusion} was previously published in \cref{harvey2023visual}; \cref{ch:fdm} is based on \cref{harvey2022flexible}; \cref{ch:tddm} is based on \cref{campbell2024trans}; and \cref{ch:cigcvae} is based on \cref{harvey2021conditional}. \Cref{ch:conclusion} contains some text from each of \cref{harvey2022flexible,harvey2021conditional}.

We also include substantial material from the following publication, for which William Harvey is second author:
\begin{enumerate}
    \setcounter{enumi}{3}
    \item \bibentry{campbell2024trans} \label{campbell2024trans}
\end{enumerate}
\Cref{ch:tddm} is based on \cref{campbell2024trans}. \Cref{ch:conclusion} contains some text from \cref{campbell2024trans}. See the final paragraph of this section for a list of description of each author's contributions.

We additionally include one figure (\cref{fig:sudoku-panel}), but no core contributions, from the following publication for which William Harvey is second author:
\begin{enumerate}
    \setcounter{enumi}{4}
    \item \bibentry{weilbach2023graphically} \label{weilbach2023graphically}
\end{enumerate}

We finally acknowledge \citeauthor{sheynin2023emu} and \citeauthor{betker2023improving} for permitting the reproduction of the images shown in \cref{fig:dalle-3-samples,fig:emu-edit-samples}. All text/figures in parts of this dissertation not mentioned here are written/created by William Harvey (WH) and are previously unpublished.

% The text in \cref{ch:flexible-diffusion} is previously unpublished and written by WH but the example and corresponding figure were published in \cref{weilbach2023graphically}.  \Cref{ch:conclusion} contains a mixture of unpublished text written by WH and text edited from \cref{harvey2022flexible,campbell2024trans,harvey2021conditional}. 

The specific contributions of each author to the papers listed above are as follows:
\begin{enumerate}
    \item WH conceptualised the idea, implemented it and ran experiments. WH wrote the paper and Frank Wood (FW) assisted with improving the presentation.
    \item WH and Christian Weilbach (CW) jointly conceptualized the method. WH implemented an initial prototype. WH and Saeid Naderiparizi (SN) jointly improved the implementation and ran experiments. CW and Vaden Masrani (VM) both also assisted with running experiments. WH, SN, VM, CW and FW all helped to workshop the presentation. WH did the majority of the writing. SN and VM also contributed significantly to the writing and presentation.
    \item WH conceptualised and implemented an initial prototype. WH and SN improved the implementation together and both ran experiments together.  SN implemented our baselines. WH, SN, and FW workshopped the presentation together. WH did the majority of the writing, while SN and FW each did some writing. WH implemented our demonstration of ``Inpainting for Bayesian Experimental Design''.
    \item Andrew Campbell (AC), WH, and CW jointly conceptualised the method. AC came up with an initial formalisation of the mathematical framework. AC and WH created two implementations, for molecular and video data respectively. AC, WH, and CW iteratively ran experiments and refined the method and proposed applications. AC and Valentin De Bortoli (VDB) produced the final formalisation of the framework. Tom Rainforth (TR) and Arnaud Doucet (AD) provided advice on the method, formalisation, applications, and presentation throughout the process.
    \item CW conceptualised the idea. WH and CW jointly came up with the methodology, implemented the method, and ran experiments. WH and CW contributed most of the writing. FW contributed some writing and assisted with workshopping the presentation.  
\end{enumerate}